Charged pion emission from neutron-rich heavy ion collisions for studies on the symmetry energy

### 2020-07-17 MSU-thesis-estee.tex
Almost final draft

% This is standard fontenc/inputenc for pdflatex
% If you use LuaLaTeX or XeLaTeX you should replace this with the fontspec package
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage{cleveref}
\usepackage{overpic}
\usepackage{contour}
\usepackage{color}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{rotating}
\usepackage{mathtools}
\usepackage{comment}\renewcommand{\arraystretch}{1.5}
\usepackage[version=3]{mhchem}

\graphicspath{{images/spacecharge/} {images/} {images/mccomparison/}  {images/dataAnalysis_2/} {images/dynamicrange/} {images/dataAnalysis/} {images/experiment/} {images/results/}  {images/intro/} }

@@

%
% Load any extra packages here
%
% You must specify the title of your thesis, your name, the field of study (not department), and the year
\title{Constraining the High Density Nuclear Symmetry Energy with Pions}
\author{Justin Estee}
\fieldofstudy{Physics} % This should be in sentence case
\date{2020}

% If you want a dedication page, specify the text of the dedication here and uncomment the next command.
%
\dedication{This thesis is dedicated to someone.}
%
%User defined commands
\newcommand{\spirit}{S$\pi$RIT }
\newcommand{\numerr}[3]{ \num[parse-numbers=false]{#1 \pm #2 (sys.) \pm #3 (stat.)}  }
%\newcommand{\numerr}[3]{ \num[input-close-uncertainty]{#1 \pm #2 (sys.) \pm #3 (stat.)}  }

@@

% All the stuff before your actual chapters is called the front matter
\frontmatter
% First make the title page
\maketitlepage
% Next make the abstract
\begin{abstract}
% Your abstract goes here.  Master's 1 page max. PhD 2 page max.
\end{abstract}

@@

% If you have a dedication page, uncomment the next command to print the dedication page
%
%\makededicationpage
%
\clearpage
% Your Acknowledgements are formatted like a chapter, but with no number
\chapter*{Acknowledgements}
\DoubleSpacing % Acknowledgements should be double spaced
Your acknowledgements here.
%
\clearpage
% We need to turn single spacing back on for the contents/figures/tables lists
\SingleSpacing
\tableofcontents* % table of contents will not be listed in the TOC
\clearpage
\listoftables % comment this out if you have no tables
\clearpage
\listoffigures % comment this out if you have no figures
%
% If you have a list of abbreviations/symbols it would go here preceded by a \clearpage
% See the class documentation and the Memoir manual for how to create other lists 
%
\mainmatter
%
% The next line removes the dots in chapter headings in the TOC
% May violate thesis office rules
%\addtocontents{toc}{\protect\renewcommand{\protect\cftchapterdotsep} {\cftnodots}}

\include{chapters/ch_intro}
%\include{chapters/ch_theory}
%\include{chapters/ch_experiment}
%\include{chapters/ch_data_analysis}
%\include{chapters/ch_data_analysis_2}
\include{chapters/ch_results}

>>>>

% This is standard fontenc/inputenc for pdflatex
% If you use LuaLaTeX or XeLaTeX you should replace this with the fontspec package
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage{overpic}
\usepackage{contour}
\usepackage{color}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{rotating}
\usepackage{mathtools}
\usepackage{appendix}
\usepackage{comment}\renewcommand{\arraystretch}{1.5}
\usepackage[version=3]{mhchem}

\graphicspath{{images/spacecharge/} {images/} {images/mccomparison/}  {images/dataAnalysis_2/} {images/dynamicrange/} {images/dataAnalysis/} {images/experiment/} {images/results/}  {images/intro/} {images/appendix/} }

@@

%
% Load any extra packages here
%
% You must specify the title of your thesis, your name, the field of study (not department), and the year
%\title{Constraining the High Density Nuclear Symmetry Energy with Pions}

%\title{Pion Emission from Neutron Rich Heavy Ion Collisions in an Effort to constrain the symmetry energy}

\title{Charged pion emission from neutron-rich heavy ion collisions for studies on the symmetry energy}

\author{Justin Brian Estee}
\fieldofstudy{Physics} % This should be in sentence case
\date{2020}

% If you want a dedication page, specify the text of the dedication here and uncomment the next command.
%
\dedication{}
%
%User defined commands
\newcommand{\spirit}{S$\pi$RIT }
\newcommand{\numerr}[3]{ \num[parse-numbers=false]{#1 \pm #2 (sys.) \pm #3 (stat.)}  }
%\newcommand{\numerr}[3]{ \num[input-close-uncertainty]{#1 \pm #2 (sys.) \pm #3 (stat.)}  }

@@

% All the stuff before your actual chapters is called the front matter
\frontmatter
% First make the title page
\maketitlepage
% Next make the abstract
\begin{abstract}
% Your abstract goes here.  Master's 1 page max. PhD 2 page max.

Recent astronomical measurements of neutron star mergers have reinvigorated questions about the nature of dense matter in the universe. The densities reached in these stars range from normal nuclear densities $\rho_o$ to upwards of $4\rho_o$ or even $9\rho_o$ in some models, where $\rho_o = \SI{1.7e14}{\gram\per\centi\metre\cubed}$; representing extremely dense, neutrons rich matter. The symmetry energy describes the difference in binding energy between pure neutron matter and symmetric matter with equal number of neutrons and protons. It is the density dependence of the symmetry energy that is related to the outward pressure, which supports the star, preventing it from collapsing under its own gravitation force. For years, theory and experiment  have combined to progressively constrain the Equation of State (EoS) of symmetric matter, and it was only in the last decade that the density dependence of pure neutron matter, the symmetry energy, had been constrained at normal nuclear densities and lower, $\rho < \rho_o$. Still large uncertainties, and discrepancies on preliminary constraints,  still remain at densities around 2$\rho_o$ which are critical importance to the understanding the properties of neutron stars.

Besides measuring neutrons stars directly, the laboratory provides the only way to study dense nuclear matter at different densities and proton-neutron asymmetries. It has been shown that charged pions are produced from nucleon-nucleon collisions through the short-lived intermediate $\Delta$ resonances, in the early, dense parts of Heavy Ion Collisions (HICs). Yet, existing pion experimental data tailored to answering questions about the symmetry energy was lacking and suffers from unknown systematic errors. this prompted the construction of the \spirit Time Projection Chamber (TPC) which was constructed to measure charged pions produced in neutron-rich HICs with a high efficiency. This Dissertation encompasses the first pion spectra from the most neutron-rich ($\tin{132}{124}$), and neutron-poor ($\tin{108}{112}$) systems measured at \SI{270}{\MeVA} taken at the RIBF facility at RIKEN. The total pion ratio was measured to within 4\% and the pion spectra was measured with a high degree of accuracy, marking the first time pion spectra  -- at energies relevant to the symmetry energy -- have been published.  Also, we will compare the results with the most recent transport models which can simulate pion production. We will highlight some of the successes and hope to motivate a discussion in the theoretical community on how to better reproduce the pion phenomena in neutron-rich HICs, which is important for not only constraining the density dependence of the symmetry energy, but also for understanding the $\Delta$ and pion roles in neutron stars.

\end{abstract}

@@

% If you have a dedication page, uncomment the next command to print the dedication page
%
%\makededicationpage
%
\clearpage
% Your Acknowledgements are formatted like a chapter, but with no number
\begin{comment}
\chapter*{Acknowledgements}
\DoubleSpacing % Acknowledgements should be double spaced
Your acknowledgements here.

%
\clearpage
\end{comment}

% We need to turn single spacing back on for the contents/figures/tables lists
\SingleSpacing
\tableofcontents* % table of contents will not be listed in the TOC
\clearpage
\listoftables % comment this out if you have no tables
\clearpage
\listoffigures % comment this out if you have no figures
%
% If you have a list of abbreviations/symbols it would go here preceded by a \clearpage
% See the class documentation and the Memoir manual for how to create other lists 
%
\mainmatter
%
% The next line removes the dots in chapter headings in the TOC
% May violate thesis office rules
%\addtocontents{toc}{\protect\renewcommand{\protect\cftchapterdotsep} {\cftnodots}}
\chapterstyle{default}

\include{chapters/ch_intro}
%\include{chapters/ch_theory}
\include{chapters/ch_experiment}
\include{chapters/ch_data_analysis}
\include{chapters/ch_data_analysis_2}
\include{chapters/ch_results}
\include{chapters/ch_summary}

### 2020-07-17 chapters/ch_appendix.tex
Almost final draft

\begin{table}[!htb]
  \begin{center}
    \begin{tabular}{ccl}
      \hline
      System & \#runs & Run numbers \\
      \hline\hline
      \hsn & 113 & 2841, 2843, 2844, 2845, 2846, 2848, 2849, 2850, 2851, 2852, 2855, 2856, \\
      & & 2857, 2858, 2859, 2860, 2861, 2875, 2877, 2878, 2879, 2880, 2881, 2882, \\
      & & 2883, 2884, 2887, 2888, 2889, 2890, 2891, 2892, 2893, 2894, 2896, 2898, \\
      & & 2899, 2900, 2901, 2902, 2903, 2904, 2905, 2907, 2914, 2916, 2917, 2919, \\
      & & 2920, 2921, 2922, 2924, 2925, 2926, 2927, 2929, 2930, 2931, 2932, 2933, \\
      & & 2934, 2935, 2936, 2939, 2940, 2941, 2942, 2943, 2944, 2945, 2946, 2948, \\
      & & 2955, 2956, 2958, 2959, 2960, 2961, 2962, 2964, 2965, 2966, 2968, 2969, \\
      & & 2970, 2971, 2972, 2973, 2975, 2976, 2977, 2978, 2979, 2980, 2981, 2982, \\
      & & 2983, 2984, 2985, 2986, 2988, 2989, 2990, 2991, 2992, 2993, 2997, 2999, \\
      & & 3000, 3002, 3003, 3007, 3039 \\
      \hline
      \mhsn & 60 & 2542, 2543, 2544, 2546, 2547, 2548, 2552, 2553, 2554, 2555, 2556, 2557, \\
      & & 2558, 2559, 2560, 2562, 2563, 2564, 2565, 2566, 2567, 2568, 2569, 2570, \\
      & & 2571, 2572, 2573, 2574, 2575, 2578, 2579, 2580, 2581, 2582, 2583, 2584, \\
      & & 2585, 2586, 2587, 2588, 2589, 2590, 2591, 2592, 2593, 2594, 2595, 2596, \\
      & & 2597, 2598, 2599, 2600, 2601, 2617, 2618, 2619, 2620, 2621, 2622, 2623  \\
      \hline
      \mlsn & 68 & 3059, 3061, 3062, 3065, 3066, 3068, 3069, 3071, 3074, 3075, 3076, 3077, \\
      & & 3078, 3080, 3081, 3082, 3083, 3084, 3085, 3087, 3088, 3089, 3090, 3091, \\
      & & 3092, 3093, 3094, 3095, 3097, 3098, 3102, 3103, 3138, 3139, 3140, 3141, \\
      & & 3142, 3143, 3144, 3145, 3146, 3148, 3149, 3150, 3151, 3152, 3153, 3154, \\
      & & 3155, 3156, 3157, 3158, 3159, 3165, 3166, 3167, 3168, 3169, 3170, 3171, \\
      & & 3172, 3177, 3179, 3180, 3181, 3182, 3183, 3184 \\
      \hline
      \lsn & 85 & 2272, 2273, 2274, 2275, 2276, 2283, 2284, 2285, 2286, 2288, 2289, 2291, \\
      & & 2310, 2311, 2314, 2315, 2320, 2322, 2323, 2324, 2325, 2331, 2332, 2333, \\
      & & 2334, 2335, 2336, 2337, 2340, 2341, 2362, 2363, 2368, 2369, 2370, 2371, \\
      & & 2372, 2373, 2374, 2375, 2378, 2379, 2380, 2381, 2382, 2383, 2384, 2385, \\
      & & 2386, 2387, 2388, 2389, 2391, 2392, 2393, 2394, 2395, 2396, 2397, 2398, \\
      & & 2399, 2400, 2401, 2402, 2429, 2432, 2433, 2434, 2437, 2438, 2439, 2440, \\
      & & 2442, 2453, 2461, 2462, 2463, 2501, 2502, 2503, 2505, 2506, 2507, 2508, \\
      & & 2509 \\
      \hline
    \end{tabular}
    \caption{List of runs for the analysis. \label{tb:runList}}
  \end{center}
\end{table}

@@

We employ a variety of track quality cuts described in Section~\ref{sec:qualitycut} to reduce the contamination from poorly reconstructed tracks which contribute to the background in the PID spectra. The best set of values were found for all the data sets which include the charged particle multiplicity of the event, the distance of closest approach to the vertex (DOCA), and the minimum number of clusters required for a track to be analyzed. By varying the cuts in both the data and in our MC efficiency calculations, we can evaluate the analysis to see how stable are the results with respect to these cuts. It is intuitive that while varying the cuts we will find that as the cut get tighter, and less data is included, the statistical error bar will increase. This problem is mitigated by looking at the uncorrelated error described in \cite{dataAnalysis}. Here if the systematic trend of the observable is much larger than the statistical error bars on the default cuts, with respects to the uncorrelated error bars, then there truly is some systematic trend that either exists due to physics or some error or miscalculation in the analysis. It is usually not recommended to estimated systematic error bars using this method, but sometimes it is the only way.

Here we will discuss the particular analysis using the total pion ratio as the example. Figure~\ref{fig:totalRatioError} shows the total pion ratio as a function of several cut variations, where one cut is varied and all the others are held constant. Three cuts are of particular interest, the event multiplicity, DOCA, and the number of cluster cut.

@@

\begin{figure}[!htb]
     \centering
     \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_NoEC_pimTotal}
         \caption{No efficiency correction.}
         \label{fig:pim_multCutVar_NoEC}
     \end{subfigure}
     \hfill
    \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_pimTotal}
         \caption{Efficiency corrected.}
         \label{fig:pim_multCutVar}
     \end{subfigure}
     \hfill
\caption{Y($\pi^-$) when varying the ${}^{132}$Sn charged particle multiplicity cut. }
\label{fig:pim_multCutVar}
\end{figure}

@@

\begin{figure}[!htb]
     \centering
     \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_NoEC_pipTotal}
         \caption{No efficiency correction.}
         \label{fig:pip_multCutVar_NoEC}
     \end{subfigure}
     \hfill
    \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_pipTotal}
         \caption{Efficiency corrected.}
         \label{fig:pip_multCutVar}
     \end{subfigure}
     \hfill
\caption{Y($\pi^+$) when varying the ${}^{132}$Sn charged particle multiplicity cut. }
\label{fig:pip_multCutVar}
\end{figure}

@@

\begin{figure}[!htb]
     \centering
     \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_NoEC_TotalRatio}
         \caption{No efficiency correction.}
         \label{fig:ratio_multCutVar_NoEC}
     \end{subfigure}
     \hfill
    \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_TotalRatio}
         \caption{Efficiency corrected.}
         \label{fig:ratio_multCutVar}
     \end{subfigure}
     \hfill
\caption{Y($\pi^+$)/Y($\pi^-$) when varying the ${}^{132}$Sn charged particle multiplicity cut. }
\label{fig:ratio_multCutVar}
\end{figure}

>>>>

\begin{table}[!htb]
  \begin{center}
    \begin{tabular}{ccl}
      \hline
      System & \#runs & Run numbers \\
      \hline\hline
      \hsn & 113 & 2841, 2843, 2844, 2845, 2846, 2848, 2849, 2850, 2851, 2852, 2855, 2856, \\
      & & 2857, 2858, 2859, 2860, 2861, 2875, 2877, 2878, 2879, 2880, 2881, 2882, \\
      & & 2883, 2884, 2887, 2888, 2889, 2890, 2891, 2892, 2893, 2894, 2896, 2898, \\
      & & 2899, 2900, 2901, 2902, 2903, 2904, 2905, 2907, 2914, 2916, 2917, 2919, \\
      & & 2920, 2921, 2922, 2924, 2925, 2926, 2927, 2929, 2930, 2931, 2932, 2933, \\
      & & 2934, 2935, 2936, 2939, 2940, 2941, 2942, 2943, 2944, 2945, 2946, 2948, \\
      & & 2955, 2956, 2958, 2959, 2960, 2961, 2962, 2964, 2965, 2966, 2968, 2969, \\
      & & 2970, 2971, 2972, 2973, 2975, 2976, 2977, 2978, 2979, 2980, 2981, 2982, \\
      & & 2983, 2984, 2985, 2986, 2988, 2989, 2990, 2991, 2992, 2993, 2997, 2999, \\
      & & 3000, 3002, 3003, 3007, 3039 \\
      \hline
      \mhsn & 60 & 2542, 2543, 2544, 2546, 2547, 2548, 2552, 2553, 2554, 2555, 2556, 2557, \\
      & & 2558, 2559, 2560, 2562, 2563, 2564, 2565, 2566, 2567, 2568, 2569, 2570, \\
      & & 2571, 2572, 2573, 2574, 2575, 2578, 2579, 2580, 2581, 2582, 2583, 2584, \\
      & & 2585, 2586, 2587, 2588, 2589, 2590, 2591, 2592, 2593, 2594, 2595, 2596, \\
      & & 2597, 2598, 2599, 2600, 2601, 2617, 2618, 2619, 2620, 2621, 2622, 2623  \\
      \hline
      \mlsn & 68 & 3059, 3061, 3062, 3065, 3066, 3068, 3069, 3071, 3074, 3075, 3076, 3077, \\
      & & 3078, 3080, 3081, 3082, 3083, 3084, 3085, 3087, 3088, 3089, 3090, 3091, \\
      & & 3092, 3093, 3094, 3095, 3097, 3098, 3102, 3103, 3138, 3139, 3140, 3141, \\
      & & 3142, 3143, 3144, 3145, 3146, 3148, 3149, 3150, 3151, 3152, 3153, 3154, \\
      & & 3155, 3156, 3157, 3158, 3159, 3165, 3166, 3167, 3168, 3169, 3170, 3171, \\
      & & 3172, 3177, 3179, 3180, 3181, 3182, 3183, 3184 \\
      \hline
      \lsn & 85 & 2272, 2273, 2274, 2275, 2276, 2283, 2284, 2285, 2286, 2288, 2289, 2291, \\
      & & 2310, 2311, 2314, 2315, 2320, 2322, 2323, 2324, 2325, 2331, 2332, 2333, \\
      & & 2334, 2335, 2336, 2337, 2340, 2341, 2362, 2363, 2368, 2369, 2370, 2371, \\
      & & 2372, 2373, 2374, 2375, 2378, 2379, 2380, 2381, 2382, 2383, 2384, 2385, \\
      & & 2386, 2387, 2388, 2389, 2391, 2392, 2393, 2394, 2395, 2396, 2397, 2398, \\
      & & 2399, 2400, 2401, 2402, 2429, 2432, 2433, 2434, 2437, 2438, 2439, 2440, \\
      & & 2442, 2453, 2461, 2462, 2463, 2501, 2502, 2503, 2505, 2506, 2507, 2508, \\
      & & 2509 \\
      \hline
    \end{tabular}
    \caption{List of all run numbers used in this analysis. \label{tb:runList}}
  \end{center}
\end{table}

@@

We employ a variety of track quality cuts described in Section~\ref{sec:qualitycut} to reduce the contamination from poorly reconstructed tracks which contribute to the background in the PID spectra. The best set of values were found for all the data sets which include the charged particle multiplicity of the event, the distance of closest approach to the vertex (DOCA), and the minimum number of clusters required for a track to be analyzed. By varying the cuts in both the data and in our MC efficiency calculations, we can evaluate the analysis to see how stable are the results with respect to these cuts. It is intuitive that while varying the cuts we will find that as the cut get tighter, and less data is included, the statistical error bar will increase. This problem is mitigated by looking at the uncorrelated error described in \cite{dataAnalysis}. Here if the systematic trend of the observable is much larger than the statistical error bars on the default cuts, with respects to the uncorrelated error bars, then there truly is some systematic trend that either exists due to physics or some error or miscalculation in the analysis. It is usually not recommended to estimated systematic error bars using this method, but sometimes it is the only way. Here we will discuss the particular analysis using the $\pi^-$ yield as an example and then discuss the $\pi^+$ the total pion ratio, and the spectral pion ratios. The three cuts used in our results are the event multiplicity, DOCA, and the number of cluster cut.

@@

\begin{figure}[!htb]
     \centering
     \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_NoEC_pimTotal}
         \caption{No efficiency correction.}
         \label{fig:pim_multCutVar_NoEC}
     \end{subfigure}
     \hfill
    \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_pimTotal}
         \caption{Efficiency corrected.}
         \label{fig:pim_multCutVar_EC}
     \end{subfigure}
     \hfill
\caption{Y($\pi^-$) when varying the ${}^{132}$Sn charged particle multiplicity cut. }
\label{fig:pim_multCutVar}
\end{figure}

@@

\begin{figure}[!htb]
     \centering
     \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_NoEC_pipTotal}
         \caption{No efficiency correction.}
         \label{fig:pip_multCutVar_NoEC}
     \end{subfigure}
     \hfill
    \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_pipTotal}
         \caption{Efficiency corrected.}
         \label{fig:pip_multCutVar_EC}
     \end{subfigure}
     \hfill
\caption{Y($\pi^+$) when varying the ${}^{132}$Sn charged particle multiplicity cut. }
\label{fig:pip_multCutVar}
\end{figure}

@@

\begin{figure}[!htb]
     \centering
     \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_NoEC_TotalRatio}
         \caption{No efficiency correction.}
         \label{fig:ratio_multCutVar_NoEC}
     \end{subfigure}
     \hfill
    \begin{subfigure}[b]{\textwidth}
         \centering
         \includegraphics[width=\textwidth]{132Sn_Multiplicity_TotalRatio}
         \caption{Efficiency corrected.}
         \label{fig:ratio_multCutVar_EC}
     \end{subfigure}
     \hfill
\caption{Y($\pi^+$)/Y($\pi^-$) when varying the ${}^{132}$Sn charged particle multiplicity cut. }
\label{fig:ratio_multCutVar}
\end{figure}
